Statistical framework for whole-genome bisulfite-sequencing data

### 2014-12-07 Rmarkdown/wgbs_statistical_framework.Rmd
TODOs for wgbs_statistical_framework

Restricting our attention to CpGs, Figure \ref{fig:EPISCOPE_beta_distribution} shows the genome-wide distributions of $\beta$-values, that is, the column-wise summaries, for each EPISCOPE sample. What is immediately clear is that these distributions are bimodal. Figure \ref{fig:EPISCOPE_beta_distribution_CGI_stratified} shows that this bimodality is largely driven by the fact that CpGs in CpG islands are mostly unmethylated whereas those outside of CpG islands are generally methylated. Intermediate $\beta$-values, those between $0.2-0.8$, are evidence of heterogeneity.

Both Figures \ref{fig:EPISCOPE_beta_distribution} and  \ref{fig:EPISCOPE_beta_distribution_CGI_stratified} also reveal that the genome-wide distributions of $\beta$-values for a given tissue are very similar between the three individuals. The three adipose-derived tissues, _SA_, _VA_ and _VAT_, have similar genome-wide profiles while the blood-derived tissue, _BUF_, has more intermediate $\beta$-values, potentially indicating more cellular heterogeneity.

```{r EPISCOPE_genome_wide_beta-value_distributions, eval = FALSE, echo = FALSE}
# Run on one of the unices
setwd("/wehisan/home/allstaff/h/hickey/methylation_m-tuples/analyses/processed_data/EPISCOPE")
library(MethylationTuples)
library(ggplot2)
library(dplyr)
library(tidyr)
thesis_theme <- theme_classic(base_size = 20)
thesis_theme <- theme_bw(base_size = 20)

x <- readRDS("EPISCOPE_1_tuples_strand_collapsed.rds")

cgi <- read.table("~/methylation_m-tuples/CGI/model-based-cpg-islands-hg19.txt",
                  header = TRUE, stringsAsFactors = FALSE)
cgi <- GRanges(seqnames = cgi$chr, IRanges(cgi$start, cgi$end))
cgi_status <- overlapsAny(x, cgi)

beta <- methLevel(x, min_cov = 10L)

cn <- colnames(beta)
beta <- data.frame(Individual = rep(substr(cn, 1, 3), each = nrow(beta)),
                   Tissue = rep(substr(cn, 4, 6), each = nrow(beta)),
                   beta = as.vector(beta),
                   CGI = rep(cgi_status, times = ncol(beta)),
                   stringsAsFactors = FALSE)
beta <- tbl_df(beta)

#a <- beta[sample(nrow(beta), 100000), ]

g <- ggplot(aes(x = beta), data = beta) +
  geom_density(size = 1.5) +
  facet_grid(Tissue ~ Individual, labeller = label_both) +
  thesis_theme + scale_colour_brewer(palette = "Dark2") +
  ggtitle(expression(paste("EPISCOPE: Distribution of ", beta, "-values", sep = ""))) +
  xlab(expression(paste(beta, "-values", sep = "")))
ggsave("EPISCOPE_beta_distribution.pdf", g, height = 9, width = 16)

g <- ggplot(aes(x = beta), data = beta) +
  geom_density(aes(colour = CGI), size = 2) +
  facet_grid(Tissue ~ Individual, labeller = label_both) +
  thesis_theme + scale_colour_brewer(palette = "Dark2") +
  ggtitle(expression(paste("EPISCOPE: Distribution of ", beta, "-values", sep = ""))) +
  xlab(expression(paste(beta, "-values", sep = "")))
ggsave("EPISCOPE_beta_distribution_CGI_stratified.pdf", g, height = 9,
       width = 16)
```

@@

```{r EPISCOPE_strand_specific_beta-value_distributions, eval = FALSE, echo = FALSE}
# Run on one of the unices
setwd("/wehisan/home/allstaff/h/hickey/methylation_m-tuples/analyses/processed_data/EPISCOPE")
library(MethylationTuples)
library(ggplot2)
library(dplyr)
library(tidyr)
library(RColorBrewer)

thesis_theme <- theme_classic(base_size = 20)
thesis_theme <- theme_bw(base_size = 20)

y <- readRDS("EPISCOPE_1_tuples.rds")
y_fwd <- y[strand(y) == "+"]
y_rev <- y[strand(y) == "-"]

ol <- findOverlaps(unstrand(y_fwd), shift(unstrand(y_rev), -1L), type = "equal")
beta_fwd <- methLevel(y_fwd, min_cov = 5L)[queryHits(ol), ]
beta_rev <- methLevel(y_rev, min_cov = 5L)[subjectHits(ol), ]

cors <- matrix(unlist(mclapply(1:10, function(min_cov, y_fwd, y_rev, ol) {
  beta_fwd <- methLevel(y_fwd, min_cov = min_cov)[queryHits(ol), ]
  beta_rev <- methLevel(y_rev, min_cov = min_cov)[subjectHits(ol), ]
  sapply(seq_len(ncol(y)), function(i, x, y) {
    cor(x[, i], y[, i], use = "complete.obs")
    }, x = beta_fwd, y = beta_rev)
  }, y_fwd = y_fwd, y_rev = y_rev, ol = ol, mc.cores = 10)), ncol = 10)

beta_pairs <- data.frame(Individual = rep(substr(cn, 1, 3),
                                          each = nrow(beta_fwd)),
                         Tissue = rep(substr(cn, 4, 6), each = nrow(beta_fwd)),
                         beta_fwd = as.vector(beta_fwd),
                         beta_rev = as.vector(beta_rev),
                         stringsAsFactors = FALSE)
beta_pairs <- tbl_df(beta_pairs)

beta_pairs %>% group_by(Individual, Tissue) %>%
  summarise(cor(beta_fwd, beta_rev, use = "complete.obs"))

mypar <- function(a = 1, b = 1, brewer.n = 8, brewer.name = "Dark2",
                  cex.lab = 2, cex.main = 2, cex.axis = 1.5,
                  mar = c(5.1, 5.1, 3.5, 2.1), mgp = c(3, 1, 0), ...) {
    par(mar = mar, mgp = mgp, cex.lab = cex.lab, cex.main = cex.main,
        cex.axis = cex.axis)
    par(mfrow = c(a, b), ...)
    palette(brewer.pal(brewer.n, brewer.name))
}
pdf("EPISCOPE_beta_correlations_across_strand.pdf", width = 16, height = 9)
mypar(4, 3)
sapply(seq_len(ncol(beta_fwd)), function(i, x, y, ns) {
  smoothScatter(x[, i], y[, i],
                xlab = expression(beta["+"]),
                ylab = expression(beta["-"]),
                main = paste0(ns[i], ": cor = ", round(cor(x[, i], y[, i],
                                                        use = "complete.obs"),
                                                    2)))
  }, x = beta_fwd, y = beta_rev, ns = colnames(beta_fwd))
dev.off()
```

>>>>

Restricting our attention to CpGs, __FIGURES__ show the kernel density estimates of the genome-wide distributions of $\beta$-values, that is, the column-wise summaries, for each sample of the EPISCOPE, Lister, Seisenberer and Ziller datasets, respectively. __FIGURES__ show the same data but with the $\beta$-values grouped into $0.01$-width bins and plotted against the percentages of CpGs that fall into each bin.

What is immediately clear is that these distributions are bimodal. The exception is the _E16.5\_male\_1_ sample from the Seisenberger data which is hypomethylated and with an enormous number of CpGs that have intermediate methylation levels. The _E16.5\_male\_1_ sample is progenitor germ cells from a pool of embryonic day 16.5 male mice. Between days E6.5 and E13.5, the mouse progenitor germ cells undergo global demethylation and it is only from day E16.5 onwards that they begin to be _de novo_ methylated \cite{Seisenberger:2012ko}.

The only samples with significant intermediate methylation are either somatic cell lines --- __LIST SAMPLES__ --- or cancer cells lines or tissue --- __LIST SAMPLES__. __EXPLANATION FOR CANCER__. __EPISCOPE DATA SHOWS ITS NOT SOMATIC CELLS BUT CELL LINES THAT GET INTERMEDIATE METHYLATION__. __HYPOTHESIS FOR SOMATIC CELL LINES__

- Most samples are hypermethylated

__ALL THE GENOME-WIDE FIGURES__

__CGI FIGURES__

@@

__TODO: Report tables of strand-correlations with different coverages__
